Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-23 agujournaltemplate.tex
Worked over Methods Section

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the community is still unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.

@@

    %%%%%%% REASSEMBLE INTRO ABOVE HERE %%%%%%%%%%%%%

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.

@@

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (more than three months) in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).

    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024). 
    %The term "Atlantic Multidecadal Oscillation" (AMO) describes the large-scale fluctuations in Atlantic sea surface temperature (SST) that occur over multiple decades \cite{kerr2000north}. This designation is intended to underscore the distinct multidecadal nature of these oceanic variations in contrast to the inter-annual variability linked with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

\subsection{Data analysis}    
    To give a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a 2-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
Variable Name & Units & Mean & Range & Std. Dev. & Data source \\
\midrule
MEI v.2 & unitless & Climate Indices provided by NOAA \\
AMO & unitless & Climate Indices provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
Evaporation & NA & ERA5 model reanalysis: evaporation \\
SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have influenced the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observe higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

@@

Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004. Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.

@@

The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community. This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}. We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

>>>>

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.

@@

@@

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program during monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts) correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e. more than three months) in coverage. Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis. Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024). This variable is referred to as "wind speed" throughout the manuscript for simplicity.

    The monthly AMO index data (10-year low-pass), which represents an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis. Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated, integrated, and then grouped by genus via the sum of interpolated counts.

\subsection{Data analysis}    
    To provide a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded due to insufficient measurements to provide a reliable yearly mean estimate.

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

    \subsubsection{Phytoplankton Community Structure Analysis}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a two-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
    % Note: Add some more info about Supplmentary Results and test runs here!

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
\toprule
Variable Name & Units  & Data source \\
\midrule
MEI v.2 & unitless & Climate index provided by NOAA \\
AMO & unitless & Climate index provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
Precipitation & m & ERA5 model reanalysis, total precipitation\\
Evaporation & m & ERA5 model reanalysis, evaporation \\
SST & °C & CTD, Average temperature measured in top 10 m\\
21° C Isotherm & m & CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

@@

Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004. Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.

@@

The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community. This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}. We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels. 
This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
